Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2022-02-09 Manuscript/Tmatrix.tex
minor corrections

To recast the BSE with a dynamical $T$-matrix kernel as an eigenvalue equation, we follow Ref.~\onlinecite{Strinati_1980} (see also Ref.~\onlinecite{Loos_2020e}) and start from the equation for the BSE amplitude 
\begin{equation}
\label{eq:Strinati_1}
	\chi_S(1,1') = \int d3d4d5d6L_0(1,4,1',3)\Xi(3,5,4,6) \chi_S(6,5)
\end{equation}
where $L_0$ is given by Eq.~\eqref{eq:L0} and the $T$-matrix kernel is
\begin{align}
\label{eq:T-matrix_kernel}
	\Xi(3,5,4,6) & =  i\frac{\delta \Sigma(3,4)}{\delta G(6,5)}\approx -T(3,5,4,6)
\end{align}
The so-called $T$-matrix self-energy $\Sigma$ is given by Eq.~\eqref{eq:SigGT} with \cite{Martin_2016,Romaniello_2012}
\begin{equation}
\label{eq:T-matrix}
	T(3,8,4,7) =
	- v_c(3,8)\delta(3,4)\delta(7,8)
	+ v_c(3,8)\delta(4,8)\delta(3,7)
	+ i \int d1'd2' v_c(3,8)G(3,1')G(8,2')T(1',2',4,7)
\end{equation}  
where \titou{$v_c$ is the bare Coulomb operator} and we neglect the functional derivative $\delta T/\delta G$ in the kernel $\Xi$.
Making the time dependence of Eq.~\eqref{eq:Strinati_1} explicit and defining $T(3,5,4,6)=- \delta(\tau^+_{35})\delta(\tau^+_{64})\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})$, one gets 
\begin{equation}
\label{eqn:DYN_T}
	\chi_S(\bx_1,\bx_{1'},\tau_{11'})e^{-i\omega_S(t_1+t_{1'})/2}
	=-i\int d3d4d5d6 G(\bx_1,\bx_3;\tau_{13})G(\bx_4,\bx_{1'};\tau_{11'})
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{-i\omega_S\tau_{34}/2}
\end{equation}
[where \titou{$\tau_{ij}=t_i-t_j$} and $\tau^+_{ij}=t^+_i-t_j$ with $t^+_i=t_i+\eta$ ($\eta \to 0^+$)] and
\begin{multline}
\label{eq:T_bar}
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) 
	= v_c(\bx_3,\bx_5)\delta(\tau_{34})\delta(\bx_3,\bx_4)\delta(\bx_6,\bx_5)-v_c(\bx_3,\bx_5)\delta(\tau_{34})	\delta(\bx_3,\bx_6)\delta(\bx_4,\bx_5)
	\\
	+ i \int d7 d8 v_c(\bx_3,x_5)G(\bx_3,\bx_7;\tau_{38})G(\bx_5,\bx_8;\tau^+_{38})\Bar{T}(\bx_7,\bx_8,\bx_4,\bx_6;\tau_{84})
\end{multline}
Using the Fourier transform $G(\tau)=\int \frac{d\omega}{2\pi}G(\omega)e^{-i\omega\tau}$, changing variable from $t_3$ to $\tau_{34}$ and taking the limit $t_{1'}=t_1^+$, we have
\begin{multline}
	\chi_S(\bx_1,\bx_{1'},0^-)
	= -i\int d\bx_3 d\bx_4 d\bx_5 d\bx_6\int d\tau_{34} \int \frac{d\omega'}{2\pi}G(\bx_1,\bx_3;\omega'+\omega_S)G(\bx_4,\bx_{1'};\omega')e^{i\omega'\tau_{34}}
	\\
	 \times \Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{-i\omega_S\tau_{34}/2}
\end{multline}
Using the Lehman representation of the one-body Green's function in the quasiparticle approximation given by Eq.~\eqref{eq:G},
%$$G(x_1,x_2;\omega)=\sum_{p}\frac{\psi_p(x_1)\psi_p^*(x_2)}{\omega-\epsilon_p+ i\eta \text{sgn}(\epsilon_p-\mu)},$$ where $\psi_p$ and $\epsilon_p$ are single-particle wavefunctions and energies, respectively, and $\mu$ the chemical potential, 
and multiply the left- and right-hand sides by $(\e{a}{}-\e{i}{}-\omega_S)\int d\bx_1 d\bx^{\prime}_1 \SO{a}(\bx_1)\SO{i}(\bx_1')$, we obtain  
\begin{multline}
\label{eq:General_DYN}
	\qty(\e{a}{}-\e{i}{}-\omega_S)\int d\bx_1d\bx^{\prime}_1 \SO{a}(\bx_1)\SO{i}(\bx_1')  \chi_S(\bx_1,\bx_{1'},0^-) =
	-\int d\bx_3 d\bx_4 d\bx_5 d\bx_6\int d\tau_{34}\SO{a}(\bx_3)\SO{i}(\bx_4)
	\\
	\times \qty[\Theta(\tau_{34})e^{i\e{i}{}\tau_{34}}+\Theta(-\tau_{34})e^{i(\e{a}{}-\omega_S)\tau_{34}} ]
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{i\omega_S\tau_{34}/2}
\end{multline}
(where $\Theta$ is the Heaviside step function) using the fact that
\begin{equation}
	\Theta(\pm\tau)e^{-i\alpha \tau}=\mp\frac{1}{2\pi i}\lim_{\eta\to 0^+}\int d\omega \frac{1}{\omega-\alpha\pm i\eta}e^{-i\omega\tau}
\end{equation}

We now consider the resonant case, \ie, $\omega_S>0$, for which 
\begin{equation}
\label{Eqn:chi_tau}
	\chi_S(\bx_1,\bx_{1'},\tau_1) 
	=-e^{i\omega_S\abs{\tau_1}/2} \sum_{pq}\SO{p}(\bx_1)\SO{q}(\bx_{1'}) \mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}
  	\qty[\Theta(\tau_1)e^{-i\e{p}{}\tau_1}+\Theta(-\tau_1)e^{-i\e{q}{}\tau_1} ]
\end{equation}
  where $\hat{c}^\dagger_{p}$ and $\hat{c}_p$ are the usual creation and annihilation operators, respectively, and $\ket{N}$ and $\ket{N,S}$ are the ground state and the $S$th excited state, respectively, of the $N$-electron system.
After some algebraic steps, one gets
\begin{multline}
	-\qty(\e{a}{}-\e{i}{}-\omega_S)\mel{N}{\hat{c}^\dagger_{i}\hat{c}_a}{N,S}
	\\
    = -\sum_{pq}\mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}
    \qty{ -\frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\ \Bar{T}_{iq,pa}(\omega) e^{-i\omega\eta} \qty[\frac{1}{\omega_S-\omega+\e{q}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{p}{}-\e{a}{}+i\eta} ] }
\end{multline}
where we have defined 
\begin{equation}
	\Bar{T}_{iq,pa}(\tau_{34})=\int d\bx_3 d\bx_4 d\bx_5 d\bx_6 \SO{a}(\bx_3) \SO{i}(\bx_4)\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) \psi_p(\bx_6)\psi^*_{q}(\bx_5)
\end{equation}
Using the definition $X_{qp,S}=\mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}$, we arrive at

\begin{equation}
	\label{eq:egval}
	\qty(\e{a}{}-\e{i}{}-\omega_S) X_{ia,S}+\sum_{pq}X_{qp,S}\widetilde{T}_{iq,pa}(\omega_S)=0
\end{equation}
where the spectral representation of the dynamical $T$-matrix is
\begin{equation}
	\widetilde{T}_{iq,pa}(\omega_S)
	= \frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\Bar{T}_{iq,pa}(\omega)e^{-i\omega\eta} \left[\frac{1}{\omega_S-\omega+\e{q}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{p}{}-\e{a}{}+i\eta}\right] 
 \label{Eqn:H2preso_eigen_2_T}
\end{equation}
Equation \eqref{eq:egval} represents a non-linear eigenvalue equation to calculate the positive excitation energies of a system, which can be rewritten as
\begin{equation}
\label{eq:H2preso_eigen}
	H_{ia,qp}(\omega_S) X_{qp,S}=\omega_S X_{ia,S}
\end{equation}
with
\begin{equation}
	H_{ia,qp}(\omega_S)=(\e{a}{}-\e{i}{})\delta_{iq}\delta_{ap} + \mel{ip}{}{aq}+\widetilde{T}_{iq,pa}(\omega_S).
\end{equation}
Here $H_{ia,jb}(\omega_S)=A_{ia,jb}(\omega_S)$ and $H_{ia,bj}(\omega_S)=B_{ia,jb}(\omega_S)$.
If one drops the correlation part of $\Bar{T}$ [the first two terms on the right-hand side of Eq.~\eqref{eq:T_bar}], one ends up the usual time-dependent Hartree-Fock (TDHF) equations. 
To calculate the correlation contribution, one must employ \eqref{eq:T} in Eq.~\eqref{Eqn:H2preso_eigen_2_T}, and after integration over the frequency, it yields Eq.~\eqref{eq:dT}.

>>>>

To recast the BSE with a dynamical $T$-matrix kernel as an eigenvalue equation, we follow Ref.~\onlinecite{Strinati_1980} (see also Ref.~\onlinecite{Loos_2020e}) and start from the equation for the BSE amplitude 
\begin{equation}
\label{eq:Strinati_1}
	\chi_S(1,1') = \int d3d4d5d6L_0(1,4,1',3)\Xi(3,5,4,6) \chi_S(6,5)
\end{equation}
where $L_0$ is given by Eq.~\eqref{eq:L0} and the $T$-matrix kernel is
\begin{align}
\label{eq:T-matrix_kernel}
	\Xi(3,5,4,6) & =  i\frac{\delta \Sigma(3,4)}{\delta G(6,5)}\approx -T(3,5,4,6)
\end{align}
The so-called $T$-matrix self-energy $\Sigma$ is given by Eq.~\eqref{eq:SigGT} with \cite{Martin_2016,Romaniello_2012}
\begin{equation}
\label{eq:T-matrix}
	T(3,8,4,7) =
	- v(3,8)\delta(3,4)\delta(7,8)
	+ v(3,8)\delta(4,8)\delta(3,7)
	+ i \int d1'd2' v(3,8)G(3,1')G(8,2')T(1',2',4,7)
\end{equation}  
where $v$ is the bare Coulomb operator and we neglect the functional derivative $\delta T/\delta G$ in the kernel $\Xi$.
Making the time dependence of Eq.~\eqref{eq:Strinati_1} explicit and defining $T(3,5,4,6)=- \delta(\tau^+_{35})\delta(\tau^+_{64})\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})$, one gets 
\begin{equation}
\label{eq:DYN_T}
	\chi_S(\bx_1,\bx_{1'},\tau_{11'})e^{-i\omega_S(t_1+t_{1'})/2}
	=-i\int d3d4d5d6 G(\bx_1,\bx_3;\tau_{13})G(\bx_4,\bx_{1'};\tau_{11'})
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{-i\omega_S\tau_{34}/2}
\end{equation}
[where $\tau_{ij}=t_i-t_j$ and $\tau^+_{ij}=t^+_i-t_j$ with $t^+_i=t_i+\eta$ ($\eta \to 0^+$)] and
\begin{multline}
\label{eq:T_bar}
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) 
	= v(\bx_3,\bx_5)\delta(\tau_{34})\delta(\bx_3,\bx_4)\delta(\bx_6,\bx_5)-v(\bx_3,\bx_5)\delta(\tau_{34})	\delta(\bx_3,\bx_6)\delta(\bx_4,\bx_5)
	\\
	+ i \int d7 d8 v(\bx_3,x_5)G(\bx_3,\bx_7;\tau_{38})G(\bx_5,\bx_8;\tau^+_{38})\Bar{T}(\bx_7,\bx_8,\bx_4,\bx_6;\tau_{84})
\end{multline}
Using the Fourier transform $G(\tau)=\int \frac{d\omega}{2\pi}G(\omega)e^{-i\omega\tau}$, changing variable from $t_3$ to $\tau_{34}$ and taking the limit $t_{1'}=t_1^+$, we have
\begin{multline}
	\chi_S(\bx_1,\bx_{1'},0^-)
	= -i\int d\bx_3 d\bx_4 d\bx_5 d\bx_6\int d\tau_{34} \int \frac{d\omega'}{2\pi}G(\bx_1,\bx_3;\omega'+\omega_S)G(\bx_4,\bx_{1'};\omega')e^{i\omega'\tau_{34}}
	\\
	 \times \Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{-i\omega_S\tau_{34}/2}
\end{multline}
Using the Lehman representation of the one-body Green's function in the quasiparticle approximation given by Eq.~\eqref{eq:G},
%$$G(x_1,x_2;\omega)=\sum_{p}\frac{\psi_p(x_1)\psi_p^*(x_2)}{\omega-\epsilon_p+ i\eta \text{sgn}(\epsilon_p-\mu)},$$ where $\psi_p$ and $\epsilon_p$ are single-particle wavefunctions and energies, respectively, and $\mu$ the chemical potential, 
and multiply the left- and right-hand sides by $(\e{a}{}-\e{i}{}-\omega_S)\int d\bx_1 d\bx^{\prime}_1 \SO{a}(\bx_1)\SO{i}(\bx_1')$, we obtain  
\begin{multline}
\label{eq:General_DYN}
	\qty(\e{a}{}-\e{i}{}-\omega_S)\int d\bx_1d\bx^{\prime}_1 \SO{a}(\bx_1)\SO{i}(\bx_1')  \chi_S(\bx_1,\bx_{1'},0^-) =
	-\int d\bx_3 d\bx_4 d\bx_5 d\bx_6\int d\tau_{34}\SO{a}(\bx_3)\SO{i}(\bx_4)
	\\
	\times \qty[\Theta(\tau_{34})e^{i\e{i}{}\tau_{34}}+\Theta(-\tau_{34})e^{i(\e{a}{}-\omega_S)\tau_{34}} ]
	\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34})\chi_S(\bx_6,\bx_5;-\tau_{34})e^{i\omega_S\tau_{34}/2}
\end{multline}
(where $\Theta$ is the Heaviside step function) using the fact that
\begin{equation}
	\Theta(\pm\tau)e^{-i\alpha \tau}=\mp\frac{1}{2\pi i}\lim_{\eta\to 0^+}\int d\omega \frac{1}{\omega-\alpha\pm i\eta}e^{-i\omega\tau}
\end{equation}

We now consider the resonant case, \ie, $\omega_S>0$, for which 
\begin{equation}
\label{eq:chi_tau}
	\chi_S(\bx_1,\bx_{1'},\tau_1) 
	=-e^{i\omega_S\abs{\tau_1}/2} \sum_{pq}\SO{p}(\bx_1)\SO{q}(\bx_{1'}) \mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}
  	\qty[\Theta(\tau_1)e^{-i\e{p}{}\tau_1}+\Theta(-\tau_1)e^{-i\e{q}{}\tau_1} ]
\end{equation}
  where $\hat{c}^\dagger_{p}$ and $\hat{c}_p$ are the usual creation and annihilation operators, respectively, and $\ket{N}$ and $\ket{N,S}$ are the ground state and the $S$th excited state, respectively, of the $N$-electron system.
After some algebraic steps, one gets
\begin{multline}
	-\qty(\e{a}{}-\e{i}{}-\omega_S)\mel{N}{\hat{c}^\dagger_{i}\hat{c}_a}{N,S}
	\\
    = -\sum_{pq}\mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}
    \qty{ -\frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\ \Bar{T}_{iq,pa}(\omega) e^{-i\omega\eta} \qty[\frac{1}{\omega_S-\omega+\e{q}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{p}{}-\e{a}{}+i\eta} ] }
\end{multline}
where we have defined 
\begin{equation}
	\Bar{T}_{iq,pa}(\tau_{34})=\int d\bx_3 d\bx_4 d\bx_5 d\bx_6 \SO{a}(\bx_3) \SO{i}(\bx_4)\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) \SO{p}(\bx_6)\SO{q}(\bx_5)
\end{equation}
Using the definition $X_{qp,S}=\mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}$, we arrive at

\begin{equation}
	\label{eq:egval}
	\qty(\e{a}{}-\e{i}{}-\omega_S) X_{ia,S}+\sum_{pq}X_{qp,S}\widetilde{T}_{iq,pa}(\omega_S)=0
\end{equation}
where the spectral representation of the dynamical $T$-matrix is
\begin{equation}
	\widetilde{T}_{iq,pa}(\omega_S)
	= \frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\Bar{T}_{iq,pa}(\omega)e^{-i\omega\eta} \left[\frac{1}{\omega_S-\omega+\e{q}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{p}{}-\e{a}{}+i\eta}\right] 
 \label{eq:H2preso_eigen_2_T}
\end{equation}
Equation \eqref{eq:egval} represents a non-linear eigenvalue equation to calculate the positive excitation energies of a system, which can be rewritten as
\begin{equation}
\label{eq:H2preso_eigen}
	H_{ia,qp}(\omega_S) X_{qp,S}=\omega_S X_{ia,S}
\end{equation}
with
\begin{equation}
	H_{ia,qp}(\omega_S)=(\e{a}{}-\e{i}{})\delta_{iq}\delta_{ap} + \mel{ip}{}{aq}+\widetilde{T}_{iq,pa}(\omega_S).
\end{equation}
Here $H_{ia,jb}(\omega_S)=A_{ia,jb}(\omega_S)$ and $H_{ia,bj}(\omega_S)=B_{ia,jb}(\omega_S)$.
If one drops the correlation part of $\Bar{T}$ [the first two terms on the right-hand side of Eq.~\eqref{eq:T_bar}], one ends up the usual time-dependent Hartree-Fock (TDHF) equations. 
To calculate the correlation contribution, one must employ \eqref{eq:T} in Eq.~\eqref{eq:H2preso_eigen_2_T}, and after integration over the frequency, it yields Eq.~\eqref{eq:dT}.
